Improving instance-level constrained clustering through Dual Iterative Local Search

### 2019-06-26 Papers/DILS/paperDILS.tex
Nuevos capitulos en memoria y correcciones en papers

\begin{abstract}
Clustering has always been a powerful tool in knowledge discovery. Traditionally unsupervised, it received renewed attention when it was shown to produce better results when provided with new types of information, thus leading to a new kind of semi-supervised learning. This new information can be given in form of instance-level must-link and cannot-link constraints, in which this paper focuses on. We propose a new metaheuristic algorithm, the Dual Iterative Local Search, and prove its capability to provide quality results for the constrained clustering problem. We will compare the results obtained by this proposal to those obtained by some of the state-of-the-art algorithms on 25 datasets with incremental levels of constraint-based information, supporting our conclusions with the aid of Bayesian statistical tests.
\end{abstract}

@@

We can define partitional clustering as the task of grouping the instances of a dataset into $k$ clusters, so that new information can be extracted from them. A dataset $X$ is composed of $n$ instances, each one of them described by $d$ features. More formally, $X = \{x_1, \cdots, x_n\}$, with the $i$th instance noted as $x_i = (x_{[i,1]}, \cdots, x_{[i,u]})$. A typical clustering algorithm assigns a class label $l_i$ to each instance $x_i \in X$. As a result, we obtain the set of labels $L = \{l_1, \cdots, l_n\}$, with $l_i \in \{1, \cdots, k\}$, that effectively splits $X$ into $k$ non-overlapping clusters $c_i$ to form a partition called $C$. The criterion used to assign an instance to a given cluster is the similarity to the rest of elements in that cluster, and the dissimilarity to the rest of instances of the dataset, which can be obtained with some kind of distance measurement \cite{jain1999data}.

@@

The constrained clustering problem can be formulated in terms of optimization, so that we can apply various optimization techniques to solve it. As mentioned earlier, it is a difficult problem to solve, so metaheuristic are presented as a promising option to find quality approximate solutions. Metaheuristic algorithms are designed to explore the solution space associated with a problem guided by a fitness (heuristic) function \cite{Gendreau:2010:HM:1941310}. In this type of approaches the key is to find a good exploration-exploitation tradeoff. This is the reason why the classic Local Search (LS) heuristic algorithm derived in the Iterative Local Search (ILS) metaheuristic algorithm, which introduces periodics perturbations in the exploration of the solution space to avoid local optima \cite{lourencco2010iterated}. ILS---or variants of it---has been applied to a wide variety of problems, including the traveling salesman \cite{archetti2018iterated}, the quadratic multiple knapsack \cite{avci2017multi} or the vehicle routing problem \cite{chentli2018impact, estrada2019biased}. ILS is also very commonly used to design hybrid methods with good exploration-exploitation tradeoff, such as its combination with expediting mechanisms \cite{zohali2019reformulation} or with the success history based differential evolution algorithm \cite{zhao2019hybrid}.

In this paper we propose a new variant of the ILS method, the Dual Iterative Local Search (DILS), which introduces diversity in the exploration of the solution space in an adaptive and guided way. We apply it to the constrained clustering problem, prooving that it constitute a good approach to obtain quality results.

Regarding the organization of this paper, Section \ref{sec:background} reviews the existing knowledge concerning constrained clustering and the state-of-the-art. In Sections \ref{sec:DILS} and \ref{sec:DILS_CC} the DILS algorithm and its constrained clustering formulation will be reviewed. Sections from \ref{sec:expSetup} to \ref{sec:analisis} present the experimental setup, results and their analysis respectively. Finally, in Section \ref{sec:conclusiones} conclusions are discussed.

@@

The first adaptation of a classic clustering method for constrained clustering was proposed in \cite{wagstaff2001constrained}. It involved modifying the widely studied K-means algorithm to take into account instance-level constraints: the already known ML and CL. This method was named COP-kmeans, it introduces a modification to the assignation rule of instances to clusters of the K-means algorithm so that an instance can be assigned to a cluster only if the assignment does not violate any constraint.

In \cite{antoine2012cecm} Constrained Evidential c-means (CECM), a variant of the Evidential c-means (ECM \cite{masson2008ecm}) algorithm is proposed, within the fuzzy clustering family of methods. The particularity of this algorithm is that the membership of instances to a cluster is defined by a probabilistic belief function. This method redefines constraints from the point of view of belief functions and includes them in the cost function.

@@

An adaptation of the Biased Random-Key Genetic Algorithm (BRKGA) \cite{gonccalves2011biased}---a variant of the Random-Key genetic algorithm \cite{bean1994genetic}---for the constrained clustering problem is proposed in \cite{de2017comparison}. It uses a genetic algorithm, combined with an LS procedure and guided by a penalty-style fitness function, to obtain a partition of the input dataset. It constitutes one of the few approaches to constrained clustering from the point of view of the evolutionary computing.

@@

The Dual Iterative Local Search (DILS) is a new variant of the classic ILS method. Its goal is to perform a search in the solution space to find the solution with the best fitness function $f$ value by introducing diversity in an adaptive and guided way. While ILS works with a single individual, DILS keeps two of them $\{m_b, m_w\}$ in memory at all times, which allows it to guide diversity-introducing methods and to avoid local optima. The individual $m_b$ is the one providing the best fitness value, so $m_w$ provides the worst fitness value.

In each generation $G$ of the optimization process DILS builds a new individual by applying a recombination operator to $m_b$ and $m_w$. After that, it applies a strong mutation operator to the newly generated individual. This is the individual DILS applies the LS procedure to in order to improve its fitness value, resulting in the trail individual $m_t$. The trial individual $m_t$ replace the worst of its parents $m_w$ if it provides a better fitness value, this operation makes the acceptance criterion for DILS. Equation \ref{eq1} displays the expression for this criterion in the case of a minimization problem.

@@

Now, the process described so far will most likely fall into a local optimum, effectively stopping the exploration of the solution space. To avoid it DILS implements a reinitialization method for $m_w$ based on the difference between the two individuals it keeps in memory $\{m_b, m_w\}$, Equation \ref{eq2} shows the reinitialization criterion for a minimization problem.

@@

\noindent where RandInit$()$ is a function that returns a randomly initialized individual and $\xi \in [-1,1]$ is the parameter that controls the tolerance of the reinitialization method. This way the best individual is always preserved and therefore always takes part in the process of generating new individuals (via recombination operator) as the best parent. Figure \ref{img:DILS_diag} summarizes the DILS optimization process.

@@

It is crucial to note that in Figure \ref{img:DILS_diag} arrows in red indicate stages of the algorithm where $m_b$ and $m_w$ are reasigned to the true best and worst individual respectively. This way, at the point where the reinitialization criterion is tested, it is possible that $f(m_w)$ provides a better value than $f(m_b)$, so that $f(m_b) - f(m_w)$ can be either positive or negative.

This leads us to note the influence of the parameter $\xi$. In one hand, if we are dealing with a minimization problem and we set $\xi > 0$, then the true worst individual is only reinitializatized when the trial individual $m_t$ replaced $m_w$ and is also better than $m_b$ by a margin. On the other hand, and also for the minimization case, but this time with $\xi < 0$, we allow DILS to reinitialize the true worst individual even when $m_t$ is not better than $m_b$. It comes clear that when $\xi = 0$ the worst individual is restarted if $f(m_b) = f(m_w)$. Note that $\xi$ could take any real value, but it is not realistic to set it outside of the range $[-1,1]$, and we recommend $[-0.5, 0.5]$. To summarize, the closer $\xi$ is to 1 the more restrictive the reinitialization criterion is---in the minimization case. It is also worth noting that in early stages of the optimization process the reinitialization criterion tends to be met more often than in later stages. The reason being is that the fitness value of the best individual $f(m_b)$ never deteriorate ($f(m_{b,G}) \ge f(m_{b,G+1})$), so it become harder for $m_b$ to get better o close to it.  Algorithm \ref{alg:DILS} summarizes the overall DILS optimization process.

@@

For the recombination and mutation operators that appear in Algorithm \ref{alg:DILS}, we use the uniform recombination and the segment mutation operator respectively. The uniform recombination consists of selecting features from the best individual $m_b$ based on a given probability $p_r$ to introduce them in the resulting individual, so the rest of the features are taken from $m_w$ \cite{spears1995virtues}. The segment mutation operator consists of replacing a fixed size segment from the features of the individual for randomly generated features. The size of the segment $p_s$ must be fixed at the beginning of the optimization process.

It should be noted that the optimization process described so far is applicable to any minimization problem, although it is clear that there is a direct adaptation of DILS for maximization problems. In addition, the LS procedure used to obtain the trial individual $m_t$ must be specified for each problem, as well as numerical representation matters and termination criteria.

@@

In this section, we present an optimization scheme for the DILS algorithm when applied to constrained clustering. We will call this new approach DILS$_{CC}$. To begin with, we will use an integer-based representation for all the individuals DILS$_{CC}$ work with; this way, each individual $m$ is defined as a vector of $n$ integers---being $n$ the number of instances of the dataset. Each integer $m_i \in [0,k)$ in $m$ represent the label of the cluster that instance $x_i$ belongs to, so each individual $m$ is a solution to the clustering problem. Additionally, in order to not bias the exploration of the solution space, we will use a random initialization to initially set $m_b$ and $m_w$, this procedure will also be used when $m_w$ must be reinitialized.

@@

We also need to specify an LS procedure for DILS to be able to improve the mutant individual to obtain the trial vector $m_t$. Algorithm \ref{alg:LS} shows the LS used in DILS$_{CC}$. Note that, in order to not to bias the search, indices from vector $m$ and labels from the set of labels are randomly chosen to generate a random subset of the neighborhood of $m$. We also stablish the maximum number of neighbors $p_m$ that can be generated in each call, this is done to find a better exploration-exploitation tradeoff.

@@

Table \ref{tab:results10} shows the results for the $SC_{10}$ constraint set. We can note that DILS$_{CC}$ represents a consistent improvement over all other 5 methods. This is due to the fact that the exploitation capacity of DILS$_{CC}$ allows it to take better advantage of the information contained in the constraint set than the rest of the methods, even with the lower level of constraint-based information we work with.

@@

Results obtained with the $SC_{15}$ constraint set are presented in Table \ref{tab:results15}. We observe how DILS$_{CC}$ and COPKM are able to obtain ARI values equal to 1, meaning that the resulting partitions of the datasets perfectly match the true partitions. The fact that DILS$_{CC}$ is always able to output a partition of the datasets is worth the advantage over COPKM. It is even the case that COPKM is not able to provide an output partition while DILS finds the optimal partition or very close to the optimal partition for the same datasets.

@@

In Table \ref{tab:results15}, which shows the results obtained with the $SC_{15}$ constraint set, we can see how DILS$_{CC}$ scales the quality of its results with the number of restrictions, as well as methods such as BRKGA or COPKM, the latter being more limited in other aspects. It is also worth noting how DILS$_{CC}$ provides the better average result for each of the three constraint set analyzed in this paper.

@@

Figure \ref{fig:heatDiag} shows all six heatmaps obtained when applying the Bayesian sign test to DILS$_{CC}$ and the state-of-the-art methods. All heatmaps suggest that the test assigns a high probability to $A - B < 0$, since most triplets are presented in the lower left third of the triangle. Bearing this in mind, and considering that ARI is a measure to maximize, we can say that the Bayesian sign test provides strong evidence that DILS$_{CC}$ does represent a significant improvement over all other six methods, being te results offered by the latter non equivalent to those offered by DILS$_{CC}$ and lower in quality.

@@

In this paper we proposed the DILS heuristic method, along with its application to the SSL constrained clustering problem, which we called DILS$_{CC}$. Focusing on the instance-level ML and CL constraints, DILS$_{CC}$ has proven that heuristic techniques constitute a good approach to constrained clustering, being able to scale the quality of the results in a way directly proportional to the number of constraints.

>>>>

\begin{abstract}
Clustering has always been a powerful tool in knowledge discovery. Traditionally unsupervised, it received renewed attention when it was shown to produce better results when provided with new types of information, thus leading to a new kind of semi-supervised learning: constrained clustering. This technique is a generalization of traditional clustering that considers additional information encoded by constraints. Constraints can be given in the form of instance-level
must-link and cannot-link constraints, which this paper focuses on. We propose a new metaheuristic algorithm, the Dual Iterative Local Search, and prove its capability to produce quality results for the constrained clustering problem. We compare the results obtained by this proposal to those obtained by previous nature-inspired techniques and by some of the state-of-the-art algorithms on 25 datasets with incremental levels of constraint-based information, supporting our conclusions with the aid of Bayesian statistical tests.
\end{abstract}

@@

We can define partitional clustering as the task of grouping the instances of a dataset into $k$ clusters, so that new information can be extracted from them. A dataset $X$ is composed of $n$ instances, each one of them described by $u$ features. More formally, $X = \{x_1, \cdots, x_n\}$, with the $i$th instance noted as $x_i = (x_{[i,1]}, \cdots, x_{[i,u]})$. A typical clustering algorithm assigns a class label $l_i$ to each instance $x_i \in X$. As a result, we obtain the set of labels $L = \{l_1, \cdots, l_n\}$, with $l_i \in \{1, \cdots, k\}$, that effectively splits $X$ into $k$ non-overlapping clusters $c_i$ to form a partition called $C$. The criterion used to assign an instance to a given cluster is the similarity to the rest of elements in that cluster, and the dissimilarity to the rest of instances of the dataset, which can be obtained with some kind of distance measurement \cite{jain1999data}.

@@

The constrained clustering problem can be formulated in terms of optimization, so that we can apply various optimization techniques to solve it. As mentioned earlier, its resolution poses a tough challenge, and metaheuristics are presented as a promising option to find quality approximate solutions. Metaheuristic algorithms are designed to explore the solution space of a problem with the guidance of a fitness (heuristic) function \cite{Gendreau:2010:HM:1941310}. In this type of approaches the key is to find a good exploration-exploitation tradeoff. This is the reason why the classic Local Search (LS) heuristic algorithm derived in the Iterative Local Search (ILS) metaheuristic algorithm, which introduces periodic perturbations in the exploration of the solution space to escape local optima \cite{lourencco2010iterated}. ILS---or variants of it---has been applied to a wide variety of problems, including the traveling salesman \cite{archetti2018iterated}, the quadratic multiple knapsack \cite{avci2017multi} or the vehicle routing problem \cite{chentli2018impact, estrada2019biased}. ILS is also very commonly used in the design of hybrid methods with good exploration-exploitation tradeoff, such as its combination with expediting mechanisms \cite{zohali2019reformulation} or with the success history based differential evolution algorithm \cite{zhao2019hybrid}.

In this paper we propose a new variant of the ILS method, the Dual Iterative Local Search (DILS), which introduces diversity in the exploration of the solution space in an adaptive and guided way. We apply it to the constrained clustering problem, proving that it constitutes a competitive approach to obtain quality results.

Regarding the organization of this paper, Section \ref{sec:background} reviews the existing knowledge concerning constrained clustering and the state-of-the-art. In Sections \ref{sec:DILS} and \ref{sec:DILS_CC} the DILS algorithm and its formulation for constrained clustering will be reviewed. Sections \ref{sec:expSetup} to \ref{sec:analisis} present the experimental setup, the results and their analysis, respectively. Finally, in Section \ref{sec:conclusiones} the conclusions are discussed.

@@

The first adaptation of a classic clustering method for constrained clustering was proposed in \cite{wagstaff2001constrained}. It involved modifying the widely studied K-means algorithm to take into account instance-level constraints: the already known ML and CL. This method, named COP-kmeans, introduces a modification to the assignment rule of instances to clusters of the K-means algorithm so that an instance can be assigned to a cluster only if the assignment does not violate any constraint.

Within the fuzzy clustering family of methods, Constrained Evidential c-means (CECM), a variant of the Evidential c-means (ECM \cite{masson2008ecm}) algorithm, is proposed in \cite{antoine2012cecm}. The particularity of this algorithm is that the membership of instances to a cluster is defined by a probabilistic belief function. This method redefines constraints from the point of view of belief functions and includes them in the cost function.

@@

An adaptation of the Biased Random-Key Genetic Algorithm (BRKGA) \cite{gonccalves2011biased}---a variant of the Random-Key genetic algorithm \cite{bean1994genetic}---for the constrained clustering problem is proposed in \cite{de2017comparison}. It uses a genetic algorithm, combined with an LS procedure and guided by a penalty-style fitness function, to obtain a partition of the input dataset. It constitutes one of the few approaches to constrained clustering from the point of view of evolutionary computing.

@@

The Dual Iterative Local Search (DILS) is a new variant of the classic ILS method. Its goal is to perform a search in the solution space to find the solution with the best fitness value (given by fitness function $f$) by introducing diversity in an adaptive and guided way. While ILS works with a single individual, DILS keeps two of them $\{m_b, m_w\}$ in memory at all times, which allows it to guide diversity-inducing methods and to avoid local optima. The individual $m_b$ is the one providing the best fitness value, whereas $m_w$ provides the worst fitness value.

In each generation $G$ of the optimization process DILS builds a new individual by applying a recombination operator to $m_b$ and $m_w$. After that, it applies a strong mutation operator to the newly generated individual. This is the individual DILS applies the LS procedure to in order to improve its fitness value, resulting in the trial individual $m_t$. The trial individual $m_t$ replaces the worst of its parents, $m_w$, if it achieves a better fitness value; this operation represents the acceptance criterion for DILS. Equation \ref{eq1} displays the expression for this criterion in the case of a minimization problem.

@@

Now, the process described so far will most likely fall into a local optimum, effectively stopping the exploration of the solution space. To avoid it DILS implements a reinitialization method for $m_w$ based on the difference between the two individuals it keeps in memory ($\{m_b, m_w\}$). Equation \ref{eq2} shows the reinitialization criterion for a minimization problem.

@@

\noindent where RandInit$()$ is a function that returns a randomly initialized individual and $\xi \in [-1,1]$ is the parameter that controls the tolerance of the reinitialization method. This way the best individual is always preserved and therefore always takes part in the process of generating new individuals (via the recombination operator) as the best parent. Figure \ref{img:DILS_diag} summarizes the DILS optimization process.

@@

It is crucial to note that in Figure \ref{img:DILS_diag} arrows in red indicate stages of the algorithm where $m_b$ and $m_w$ are reassigned to the true best and worst individual respectively. This way, at the point where the reinitialization criterion is tested, it is possible for $f(m_w)$ to yield a smaller value than $f(m_b)$, so that $f(m_b) - f(m_w)$ can be either positive or negative.

This highlights the influence of the parameter $\xi$. Let us assume again a minimization problem. On one hand, if $\xi > 0$, then the true worst individual is only reinitialized when $m_t$ replaces $m_w$ and is also better than $m_b$ by a margin. On the other hand, if $\xi < 0$, we allow DILS to reinitialize the true worst individual even when $m_t$ is not better than $m_b$. It becomes clear that when $\xi = 0$ the worst individual is restarted if $f(m_b) = f(m_w)$. Note that $\xi$ could take any real value, but it is not realistic to set it outside of the range $[-1,1]$, and we recommend $[-0.5, 0.5]$. To summarize, the closer $\xi$ is to 1 the more restrictive the reinitialization criterion is---in the minimization case. It is also worth noting that in early stages of the optimization process the reinitialization criterion tends to be met more often than in later stages. The reason for this is that the fitness value of the best individual $f(m_b)$ never worsens ($f(m_{b,G}) \ge f(m_{b,G+1})$), so it becomes increasingly harder for $m_w$ to achieve similar or better scores. Algorithm \ref{alg:DILS} summarizes the overall DILS optimization process.

@@

For the recombination and mutation operators that appear in Algorithm \ref{alg:DILS}, we use the uniform recombination and the segment mutation operator respectively. The uniform recombination consists in selecting features from the best individual $m_b$ based on a given probability $p_r$ to introduce them in the resulting individual, so the rest of the features are taken from $m_w$ \cite{spears1995virtues}. The segment mutation operator consists in replacing a fixed-size segment from the features of the individual with randomly generated features. The size of the segment $p_s$ must be fixed at the beginning of the optimization process.

It should be noted that the optimization process described so far is applicable to any minimization problem, although it is clear that there is a direct adaptation of DILS for maximization problems. In addition, the LS procedure used to obtain the trial individual $m_t$ must be specified for each problem, as well as numerical representation details and termination criteria.

@@

In this section, we present an optimization scheme for the DILS algorithm when applied to constrained clustering. We will call this new approach DILS$_{CC}$. We will use an integer-based representation for all the individuals DILS$_{CC}$ work with; this way, each individual $m$ is defined as a vector of $n$ integers---with $n$ being the number of instances of the dataset. Each integer $m_i \in [0,k)$ in $m$ represents the label of the cluster that instance $x_i$ is assigned to, and therefore each individual $m$ is a solution to the clustering problem. Additionally, in order not to bias the exploration of the solution space, we will use random initialization to set $m_b$ and $m_w$; this procedure will also be used in the reinitialization of $m_w$.

@@

We also need to specify an LS procedure for DILS to be able to improve the mutant individual to obtain the trial vector $m_t$. Algorithm \ref{alg:LS} shows the LS used in DILS$_{CC}$. Note that, in order not to bias the search, indices from vector $m$ and labels from the set of labels are randomly chosen to generate a random subset of the neighborhood of $m$. We also establish the maximum number of neighbors $p_m$ that can be generated in each call; this is done to ensure a better exploration-exploitation tradeoff.

@@

Table \ref{tab:results10} shows the results for the $SC_{10}$ constraint set. It is possible to notice that DILS$_{CC}$ represents a consistent improvement over all other 5 methods. This is due to the fact that the exploitation capacity of DILS$_{CC}$ allows it to take better advantage of the information contained in the constraint set than the rest of the methods, even with the lowest level of constraint-based information we work with.

@@

The results obtained with the $SC_{15}$ constraint set are presented in Table \ref{tab:results15}. We observe how DILS$_{CC}$ and COPKM are able to obtain ARI values equal to 1, meaning that the resulting partitions of the datasets perfectly match the true partitions. The fact that DILS$_{CC}$ is always able to output a partition of the datasets is a noteworthy advantage over COPKM. Even in the cases where COPKM is unable to produce a partition, DILS often finds the optimal or a near-optimal partition.

@@

In Table \ref{tab:results15}, which shows the results obtained with the $SC_{15}$ constraint set, we can see how the quality of the results of DILS$_{CC}$ scales with the number of constraints, as well as that of methods such as BRKGA or COPKM (the latter being more limited in other aspects). It is also worth noting how DILS$_{CC}$ provides the best average ARI for each of the three constraint sets analyzed in this paper.

@@

Figure \ref{fig:heatDiag} shows all six heatmaps obtained when applying the Bayesian sign test to DILS$_{CC}$ and the state-of-the-art methods. All heatmaps suggest that the test assigns a high probability to $A - B < 0$, since most triplets are presented in the lower left third of the triangle. Bearing this in mind, and considering that ARI is a measure to maximize, we can say that the Bayesian sign test provides strong evidence that DILS$_{CC}$ does represent a significant improvement over all other six methods, being the results offered by the latter non equivalent to those offered by DILS$_{CC}$ and lower in quality.

@@

In this paper we proposed the DILS heuristic method, along with its application to the SSL constrained clustering problem, which we called DILS$_{CC}$. Focusing on instance-level ML and CL constraints, DILS$_{CC}$ has proven that heuristic techniques constitute a competitive approach to constrained clustering, being able to scale the quality of the results in a directly proportional way to the number of constraints.
